ANCOVA Comparison Simulations: Skewed, Poisson Data

### 2017-02-16 simulations/simulation_skewed.Rmd
Clean up suppl simulation files

We use two designs:

* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$. This is a constant, additive treatment effect at stratum 1, but no treatment effect at strata 2 and 3. This is a simplistic case of heterogeneous treatment effects.

With these two designs, we vary the distribution of $\varepsilon$.
In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANCOVA assumptions.
In the second case, $\varepsilon \sim t(2)$ so the errors are heavy-tailed.
Thus, there are four total simulation designs.

We compare five tests:

* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
* Freedman-Lane test: see the other Rmd document for a full description of this procedure

Throughout our simulations, we first fix $Y_0$ and stratum ID.
Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
Then, $Y_1$ is constructed using the linear data-generating process above.
We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.

@@

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title)
}

plot_pvalue_hist <- function(pvalue_mat, title){
  melt(pvalue_mat) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes(x = value, fill = Method)) +
  geom_histogram() +
  facet_wrap(~Method) +
  ggtitle(title)
}

plot_pvalue_scatter <- function(pvalue_mat, title){
  pvalue_mat %>%
  as.data.frame() %>%
  select(ANCOVA, strat = starts_with("Stratified")) %>%
  ggplot(aes(x = ANCOVA, y = strat)) + 
  geom_point() +
  xlim(0, 1) + ylim(0, 1) +
  ylab("Stratified Permutation") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  ggtitle(title)
}
```

@@

Before testing for different kinds of effects, we begin checking that the tests have the correct level.
We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
Figure \ref{fig:design_null_plots} demonstrates that this is the case.
If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.

@@

```{r design_null_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design0_power, "No Effect, Balanced Design")

plot_pvalue_hist(design0_pvalues, "No Effect, Balanced Design")
```

@@

There is no discernable difference in power between the ANCOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
Without controlling for the baseline values, the variance in $Y_1$ masks the treatment effect.

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
p1 <- plot_power_curves(design1_power, "Constant Additive Effect of 0.5")
p1
plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 0.5")
```

@@

A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
All five power curves are closer to the line passing through the origin: since the effect is only present at one stratum, it is more difficult to detect.

@@

```{r design2_plot, fig.height = 4, fig.align="center"}
p2 <- plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
p2
plot_pvalue_hist(design2_pvalues, "Effect of 0.5 in a Single Stratum")
```

@@

```{r design3_plot, fig.height = 4, fig.align="center"}
p3 <- plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
p3
plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
```

@@

```{r design_null_imbalanced_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design00_power, "No Effect, Imbalanced design")

plot_pvalue_hist(design00_pvalues, "No Effect, Imbalanced design")
```

@@

```{r design4_plot, fig.height = 4, fig.align="center"}
p4 <- plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
p4
plot_pvalue_hist(design4_pvalues, "Constant Additive Effect of 0.5 \n Imbalanced Design")
```

@@

There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
The treatment effect is on the same order of magnitude as the error variance.

@@

```{r design5_plot, fig.height = 4, fig.align="center"}
p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single stratum, Imbalanced Design")
p5 
plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single stratum, Imbalanced Design")
```

```{r final_summary, fig.align = "center"}
#plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
#do.call(grid.arrange, plots4)

powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Design" = rep("Balanced", nrow(design1_power))), 
               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design2_power)), "Design" = rep("Balanced", nrow(design2_power))), 
               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design4_power)), "Design" = rep("Imbalanced", nrow(design4_power))), 
               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))),
               design0_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design0_power)), "Design" = rep("Balanced", nrow(design0_power))),
               design00_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design00_power)), "Design" = rep("Imbalanced", nrow(design00_power))))

>>>>

We use the same simulation guidelines as detailed in the Gaussian simulation file.
We omit comments.

@@

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}
```

@@

@@

```{r design_null_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design0_power, "No Effect \n Balanced Design")
```

@@

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design1_power, "Constant Additive Effect of 0.5")

```

@@

@@

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
```

@@

```{r design3_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
```

@@

```{r design_null_imbalanced_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design00_power, "No Effect \n Imbalanced design")
```

@@

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
```

@@

@@

```{r design5_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design5_power, "Effect of 0.5 at a Single stratum \n Imbalanced Design")
```

```{r final_summary, fig.align = "center"}
powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Design" = rep("Balanced", nrow(design1_power))), 
               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design2_power)), "Design" = rep("Balanced", nrow(design2_power))), 
               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design4_power)), "Design" = rep("Imbalanced", nrow(design4_power))), 
               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))),
               design0_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design0_power)), "Design" = rep("Balanced", nrow(design0_power))),
               design00_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design00_power)), "Design" = rep("Imbalanced", nrow(design00_power))))
